### 2022-07-19 draft/draft.Rmd
reworking some of the outline/structure

```{=tex}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{remark}
\newtheorem{remark}{Remark}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
```
# Introduction

@@

Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science. 
This has resulted in various challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors. 
To illustrate this challenge, consider the typical approach to building a statistical or machine learning model. 
Data are often represented as an $n \times p$ matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ in which each row $x_i \in \mathbb{R}^p$ is a vector of $p$ features and each column is a set of $n$ feature measurements. 
An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector. 
For instance, $\phi(x)$ might first compute the distance from $x$ to one of $K$ points in $\mathbb{R}^p$ and assign $x$ to the label of the nearest point. 
Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm or Gaussian mixture models (in the case of unsupervised learning). 
However, this type of approach is incompatible with data that are represented as graphs, in which each observation is not a vector of numerical features but a set of relationships to other observations. 
Instead of feature vector $x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top$, we observe $a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}$ where each $a_{ij}$ is object $i$'s relationship to object $j$.

@@

Given a sample or dataset, a typical analysis tasks are statistical inference, or the estimation of various parameters assuming the data come from a random distribution or process. 
These estimated parameters are often then used for making predictions or deriving insights about the population. 
For example, when fitting a Gaussian mixture model, the data are first assumed to come from a mixture of Gaussians. 
The model fitting process then involves estimating the means and standard deviations of each Gaussian component, along with the mixture weights. 
The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components, as well as a prediction for which mixture a new observation belongs to. 
In order to perform a similar type of analysis for graphs, we must first define probability distributions from which such data can be sampled.

Within the scope of this dissertation, we focus primarily on unweighted and undirected graphs without self-loops. 
The adjacency matrix that describes such a graph is binary, symmetric, and hollow. 
In this setting, a plausible model is to sample each edge independently from a Bernoulli distribution, i.e., $A_{ij} \indep P_{ij}$ for some $P_{ij} \in [0, 1]$ for each $i < j$ (setting $A_{ji} = A_{ij}$ since $A$ is symmetric, and $A_{ii} = 0$ since $A$ is hollow). 
Then similar to how the edges are compiled into an adjacency matrix $A$, the edge probabilities can be compiled into an edge probability matrix $P \in [0, 1]^{n \times n}$. 
In order to provide structure

@@

### Review of Likelihood Maximization Approaches to Block Models

>>>>

```{=tex}
\setlength{\parindent}{10.0pt}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{remark}
\newtheorem{remark}{Remark}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}
```
# Introduction

@@

Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science. 
This has resulted in challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors. 
To illustrate this, consider the typical approach to building a statistical or machine learning model. 
Data are often represented as an $n \times p$ matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ in which each row $x_i \in \mathbb{R}^p$ is an observation of $p$ features and each column is a set of $n$ feature measurements. 
An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector. 
For instance, $\phi(x)$ might first compute the distance from $x$ to one of $K$ points in $\mathbb{R}^p$ and assign $x$ to the label of the nearest point. 
Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm or Gaussian mixture models (in the case of unsupervised learning). 
However, this type of approach is incompatible with data that are represented as graphs, in which each observation is not a vector of numerical features but a set of relationships to other observations. 
Instead of feature vector $x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top$, we observe $a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}$ where each $a_{ij}$ is object $i$'s relationship to object $j$.

@@

Given a sample or dataset, a typical analysis task is statistical inference, or the estimation of various parameters assuming the data come from a random distribution or process. 
These estimated parameters are often then used for making predictions or deriving insights about the population. 
For example, when fitting a Gaussian mixture model, the data are first assumed to come from a mixture of Gaussians. 
The model fitting process then involves estimating the means and standard deviations of each Gaussian component, along with the mixture weights. 
The resulting model provides insight into where each mixture component is located, how disperse each component is, and how the data are distributed between the components, as well as a prediction indicating to which mixture a new observation belongs. 
In order to perform a similar type of analysis for graphs, we must first define probability distributions from which such data can be sampled.

Within the scope of this dissertation, we focus primarily on unweighted and undirected graphs without self-loops, with a brief discussion on generalizing these methods to weighted or directed graphs. 
The adjacency matrix that describes these graphs is binary, symmetric, and hollow. 
In this setting, a plausible model is to sample each edge independently from a Bernoulli distribution, i.e., $A_{ij} \indep P_{ij}$ for some $P_{ij} \in [0, 1]$ for each $i < j$ (setting $A_{ji} = A_{ij}$ since $A$ is symmetric, and $A_{ii} = 0$ since $A$ is hollow). 
Then similar to how the edges are compiled into an adjacency matrix $A$, the edge probabilities can be compiled into an edge probability matrix $P \in [0, 1]^{n \times n}$. 
This type of graph model is defined as a *Bernoulli graph*.

\begin{definition}[Bernoulli graph]
\end{definition}

Statistical inference is not possible on a general Bernoulli graph with arbitrary edge probabilities since the number of parameters is equal to the number of observations. 
Additional structure must be introduced. 
One such structured Bernoulli graph model is the $\ER$ graph, which is defined as follows.

\begin{definition}[$\ER$ graph]
\end{definition}

@@

### The Hierarchy of Block Models

### Review of Traditional Statistical Inference Approaches to Block Model Estimation

### 2022-07-19 draft/draft.tex
reworking some of the outline/structure

\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{remark}
\newtheorem{remark}{Remark}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}

@@

Graph and network data have become increasingly widespread in various
fields including sociology, neuroscience, biostatistics, and computer
science. This has resulted in various challenges for researchers who
rely on traditional statistical and machine learning methods that are
incompatible with graph data and instead assume that the data exist as
feature vectors. To illustrate this challenge, consider the typical
approach to building a statistical or machine learning model. Data are
often represented as an \(n \times p\) matrix
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\) in which
each row \(x_i \in \mathbb{R}^p\) is a vector of \(p\) features and each
column is a set of \(n\) feature measurements. An analysis task for
these data might be to come up with a classification model
\(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical
values of each feature of a vector. For instance, \(\phi(x)\) might
first compute the distance from \(x\) to one of \(K\) points in
\(\mathbb{R}^p\) and assign \(x\) to the label of the nearest point.
Examples of this include linear discriminant analysis (in the case of
supervised learning) and Lloyd's algorithm or Gaussian mixture models
(in the case of unsupervised learning). However, this type of approach
is incompatible with data that are represented as graphs, in which each
observation is not a vector of numerical features but a set of
relationships to other observations. Instead of feature vector
\(x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top\),
we observe
\(a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}\) where
each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).

@@

Given a sample or dataset, a typical analysis tasks are statistical
inference, or the estimation of various parameters assuming the data
come from a random distribution or process. These estimated parameters
are often then used for making predictions or deriving insights about
the population. For example, when fitting a Gaussian mixture model, the
data are first assumed to come from a mixture of Gaussians. The model
fitting process then involves estimating the means and standard
deviations of each Gaussian component, along with the mixture weights.
The resulting model provides insight into where each mixture component
is located, how disperse each component is, and how the data are
distributed between the components, as well as a prediction for which
mixture a new observation belongs to. In order to perform a similar type
of analysis for graphs, we must first define probability distributions
from which such data can be sampled.

Within the scope of this dissertation, we focus primarily on unweighted
and undirected graphs without self-loops. The adjacency matrix that
describes such a graph is binary, symmetric, and hollow. In this
setting, a plausible model is to sample each edge independently from a
Bernoulli distribution, i.e., \(A_{ij} \indep P_{ij}\) for some
\(P_{ij} \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\)
since \(A\) is symmetric, and \(A_{ii} = 0\) since \(A\) is hollow).
Then similar to how the edges are compiled into an adjacency matrix
\(A\), the edge probabilities can be compiled into an edge probability
matrix \(P \in [0, 1]^{n \times n}\). In order to provide structure

@@

\hypertarget{review-of-likelihood-maximization-approaches-to-block-models}{%
\subsubsection{Review of Likelihood Maximization Approaches to Block
Models}\label{review-of-likelihood-maximization-approaches-to-block-models}}

>>>>

\setlength{\parindent}{10.0pt}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\tr}{\mathrm{Tr}}
\newcommand{\blockdiag}{\mathrm{blockdiag}}
\newcommand{\indep}{\stackrel{\mathrm{ind}}{\sim}}
\newcommand{\iid}{\stackrel{\mathrm{iid}}{\sim}}
\newcommand{\Bernoulli}{\mathrm{Bernoulli}}
\newcommand{\Betadist}{\mathrm{Beta}}
\newcommand{\BG}{\mathrm{BernoulliGraph}}
\newcommand{\Uniform}{\mathrm{Uniform}}
\newcommand{\PABM}{\mathrm{PABM}}
\newcommand{\RDPG}{\mathrm{RDPG}}
\newcommand{\GRDPG}{\mathrm{GRDPG}}
\newcommand{\Multinomial}{\mathrm{Multinomial}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{remark}
\newtheorem{remark}{Remark}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\as}{\stackrel{\mathrm{a.s.}}{\to}}
\newcommand{\ER}{\text{Erd\"{o}s-R\'{e}nyi}}

@@

Graph and network data have become increasingly widespread in various
fields including sociology, neuroscience, biostatistics, and computer
science. This has resulted in challenges for researchers who rely on
traditional statistical and machine learning methods that are
incompatible with graph data and instead assume that the data exist as
feature vectors. To illustrate this, consider the typical approach to
building a statistical or machine learning model. Data are often
represented as an \(n \times p\) matrix
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\) in which
each row \(x_i \in \mathbb{R}^p\) is an observation of \(p\) features
and each column is a set of \(n\) feature measurements. An analysis task
for these data might be to come up with a classification model
\(\phi : \mathbb{R}^p \to \{1, 2, ..., K\}\) that uses the numerical
values of each feature of a vector. For instance, \(\phi(x)\) might
first compute the distance from \(x\) to one of \(K\) points in
\(\mathbb{R}^p\) and assign \(x\) to the label of the nearest point.
Examples of this include linear discriminant analysis (in the case of
supervised learning) and Lloyd's algorithm or Gaussian mixture models
(in the case of unsupervised learning). However, this type of approach
is incompatible with data that are represented as graphs, in which each
observation is not a vector of numerical features but a set of
relationships to other observations. Instead of feature vector
\(x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top\),
we observe
\(a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}\) where
each \(a_{ij}\) is object \(i\)'s relationship to object \(j\).

@@

Given a sample or dataset, a typical analysis task is statistical
inference, or the estimation of various parameters assuming the data
come from a random distribution or process. These estimated parameters
are often then used for making predictions or deriving insights about
the population. For example, when fitting a Gaussian mixture model, the
data are first assumed to come from a mixture of Gaussians. The model
fitting process then involves estimating the means and standard
deviations of each Gaussian component, along with the mixture weights.
The resulting model provides insight into where each mixture component
is located, how disperse each component is, and how the data are
distributed between the components, as well as a prediction indicating
to which mixture a new observation belongs. In order to perform a
similar type of analysis for graphs, we must first define probability
distributions from which such data can be sampled.

Within the scope of this dissertation, we focus primarily on unweighted
and undirected graphs without self-loops, with a brief discussion on
generalizing these methods to weighted or directed graphs. The adjacency
matrix that describes these graphs is binary, symmetric, and hollow. In
this setting, a plausible model is to sample each edge independently
from a Bernoulli distribution, i.e., \(A_{ij} \indep P_{ij}\) for some
\(P_{ij} \in [0, 1]\) for each \(i < j\) (setting \(A_{ji} = A_{ij}\)
since \(A\) is symmetric, and \(A_{ii} = 0\) since \(A\) is hollow).
Then similar to how the edges are compiled into an adjacency matrix
\(A\), the edge probabilities can be compiled into an edge probability
matrix \(P \in [0, 1]^{n \times n}\). This type of graph model is
defined as a \emph{Bernoulli graph}.

\begin{definition}[Bernoulli graph]
\end{definition}

Statistical inference is not possible on a general Bernoulli graph with
arbitrary edge probabilities since the number of parameters is equal to
the number of observations. Additional structure must be introduced. One
such structured Bernoulli graph model is the \(\ER\) graph, which is
defined as follows.

\begin{definition}[$\ER$ graph]
\end{definition}

@@

\hypertarget{the-hierarchy-of-block-models}{%
\subsubsection{The Hierarchy of Block
Models}\label{the-hierarchy-of-block-models}}

\hypertarget{review-of-traditional-statistical-inference-approaches-to-block-model-estimation}{%
\subsubsection{Review of Traditional Statistical Inference Approaches to
Block Model
Estimation}\label{review-of-traditional-statistical-inference-approaches-to-block-model-estimation}}
